\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-22 paper/algorithm.tex
paper: rework algorithm and conclusion to account for new cost model

\subsection{Cost model}
\label{sec:cost-model}
We rank query plans using a cost model \[\textsc{Cost}(q) = (1 - \alpha) \times \textsc{Time}(q) + \alpha \times \frac{\Sigma_i \textsc{Loss}(q)_i}{|\textsc{Loss}(q)|}.\] $\textsc{Time}(q)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates. $\textsc{Loss}(q)$ is an estimate of the amount of error introduced by a sequence of imputation procedure. Our
cost model averages over the losses from each such operator. 
$\alpha$ is a tunable parameter that determines whether the optimizer focuses on quality or on performance. $\alpha = 1.0$ means that the optimal query should have the highest quality imputation (at the expense of performance), $\alpha=0.0$ means that the optimal query should be as fast as possible (at the expense of imputation quality).

@@

Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
At a high level, a plan cache is a mapping from a set of tables to a set of optimal plans that perform the required joins and filters over these tables.
We store multiple plans rather than a single best plan because each plan has a distinct set of dirty attributes.

\begin{figure}
\todo{formatting of plan cache semantics needs to be fixed}
\todo{I think the cache update step is not quite right, since p seems to be a single plan, but it is really a set}
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T, P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned}
                                           Q \cup \{(T,\ \argmin_{q \in P}\ \textsc{Cost}(q))\}\ \text{where} \\
                                           P = \left\{q ~|~ \begin{aligned}
                                                                     q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p) 
                                                                   \end{aligned}
                                           \right\}
                                         \end{aligned}
    % \in 
  \end{align*}
  \caption{Semantics of the plan cache.}
\end{figure}

>>>>

\subsection{Cost model}
\label{sec:cost-model}
The cost for a plan $q$ is expressed as a tuple of $\langle \textsc{Loss}(q), \textsc{Time}(q) \rangle$.
$\textsc{Time}(q)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates. $\textsc{Loss}(q)$ is an estimate of the amount of error introduced by a sequence of imputation procedure. Our
cost model averages over the losses from each such operator. We note that our loss calculation constrains $\textsc{Loss}(q)$ to $[0, 1]$, while $\textsc{Time}(q)$ can range within $(0, \infty)$.

\begin{definition}
We say that a plan $p$ with cost $\langle l, t \rangle$ dominates a plan $p'$ with cost $\langle l', t' \rangle$, if $l < l' \land t < t'$, and denote this as $\doms{p}{p'}$. 
\end{definition}

Given a query, \ProjectName{} produces a final set of possible plans

\begin{gather*}
 P= \{ p | \forall p'. \doms{p}{p'} \}
 \end{gather*}

This set outlines the best possible options for trading performance and cost for the current query, building a Pareto frontier\cite{pareto1964cours}.

In order to pick a final plan from the frontier, our model introduces a parameter $\alpha$ which encodes an upper bound on the amount of information loss that a user is willing to tolerate
relative to the minimum information loss possible given the imputation operations.

\begin{definition}
Given a final set of plans $P$, and $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Loss}(p) - \min_{p' \in P}(\textsc{Loss}(p')) \leq \alpha$. We denote this as $p^\alpha$.
We say $p$ is $\alpha$-bound optimal if $\forall p'^\alpha \in P: \textsc{Time}(p) \leq \textsc{Time}(p')$. 
\end{definition}

Given $\alpha$ and $P$, \ProjectName{} returns the $\alpha$-bound optimal plan in $P$. Note that the $\alpha$-bound optimal plan is unique, given that
$P$ only maintains dominating plans.

So in essence, $\alpha$ is a tunable parameter that determines whether the optimizer focuses on quality or on performance, as plans that
lose a lot of information through dropping are also the fastest.

$\alpha = 1.0$ means that the optimal query should have the highest performance (at the expense of quality), while $\alpha=0.0$ means that the optimal query should lose
as little information as possible (at the expense of performance).

@@

Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
At a high level, a plan cache is a mapping from a set of tables to a set of dominating plans that perform the required joins and filters over these tables.
We store multiple plans rather than a single plan because each plan has a distinct set of dirty attributes and a choice over the plan frontier
can only be made once the set is final. Details are shown in Figure~\ref{fig:semantics-plan-cache}.

\begin{figure}
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T, P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P') \} \end{aligned}\\
                                            \text{ where } \\
                                           P' &= \left\{q ~|~ \begin{aligned}
                                                                     q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p),\\
                                                                     \forall q' \in \llbracket Q[T] \rrbracket.  \doms{q} {q'} 
                                                                   \end{aligned}
                                           \right\}
  \end{align*}
  \caption{Semantics of the plan cache.}
  \label{fig:semantics-plan-cache}
\end{figure}

### 2017-02-22 paper/algorithms/plan-toplevel.tex
paper: keep refining algorithm section
paper: rework algorithm and conclusion to account for new cost model

% \begin{algorithm}
\begin{algorithmic}
  \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$, and an optional set of grouping attributes $G$ and aggregator function $A$.}
  \Ensure{An optimized query plan.}

  \Function{Plan}{$T, F, J, P, G, A$}
  \State Let $Q$ be an empty plan cache.
  \For{$t \in T$} \Comment{Add selections to the plan cache.}
  \If {$\exists \phi: (t, \phi) \in F$}
  \State $Q[\{t\}] \lhd \Call{OptFilter}{Q, t, \phi, P}$
  \Else
  \State $Q[\{t\}] \lhd \{t\}$
  \EndIf
  \EndFor

  \State $\Call{OptJoin}{Q, T, J, P}$	\Comment{Optimize joins.}
  \State $B \gets Q[T]$ \Comment{Get the best plans for all tables.}
  \If{$G \neq \emptyset$} \Comment{Add optional group \& aggregate.}
  \State $B \gets \bigcup_{q \in B} \Call{OptGroupBy}{q, G, A, C_{vis}}$
  \EndIf

  \State $B \gets \bigcup_{q \in B} \Call{OptProject}{q, P}$ \Comment{Add projections.}

  \State \Return $\argmin_{q \in B} Cost(q)$ \Comment{Return cheapest plan.}
  \EndFunction

>>>>

% \begin{algorithm}
\begin{algorithmic}[5]
  \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$,  an optional set of grouping attributes $G$ and aggregator function $A$, and a parameter $\alpha \in [0, 1]$ that expresses trade-off performance and information loss}
  \Ensure{An optimized query plan.}

  \Function{Plan}{$T, F, J, P, G, A$}
  \State $C_g \gets \bigcup_{\phi \in F} \Call{Attr}{\phi}$ \Comment{Collect relevant attributes.} \label{lst:line:attr-start}
  \State $C_g \gets \bigcup_{\psi \in J} \Call{Attr}{\psi} \cup P \cup G \cup \Call{Attr}{A} \cup C_g$ \label{lst:line:attr-end}
  \State Let $Q$ be an empty plan cache.
  \For{$t \in T$} \Comment{Add selections to the plan cache.}
  \If {$\exists \phi: (t, \phi) \in F$}
  \State $Q[\{t\}] \lhd \Call{OptFilter}{Q, t, \phi, C_g}$
  \Else
  \State $Q[\{t\}] \lhd \{t\}$
  \EndIf
  \EndFor

  \State $\Call{OptJoin}{Q, T, J, C_g}$	\Comment{Optimize joins.}
  \State $B \gets Q[T]$ \Comment{Get the best plans for all tables.}
  \If{$G \neq \emptyset$} \Comment{Add optional group \& aggregate.}
  \State $B \gets \bigcup_{q \in B} \Call{OptGroupBy}{q, G, A, C_g}$
  \EndIf

  \State $B \gets \bigcup_{q \in B} \Call{OptProject}{q, P}$ \Comment{Add projections.}
  \State \Return $p \in B$ s.t. $p$ is $\alpha$-bound optimal  \Comment{Return fastest plan satisfying $\alpha$ bound.}
  \EndFunction

### 2017-02-22 paper/conclusion.tex
paper: rework algorithm and conclusion to account for new cost model

\textbf{Cost model interpretation:} While the current system
provides a parameter ($\alpha$) to tradeoff between imputation quality and runtime,
there is no direct interpretation of its unit-less value. One natural extension to this work
is to reformulate the cost-model in an interpretable fashion.

@@

\textbf{Adaptive query planning:} Currently, the imputation is local to a given query, meaning
no information is shared across queries. An intriguing direction would be to take advantage
of imputation and execution data generated by repeated queries. Note that this goes beyond
simply caching imputation results, and instead could entail operations such as extending
intermediate results with prior imputation values to improve accuracy, or pruning out
query plans which have shown to produce low-confidence imputations.

>>>>

@@

\textbf{Adaptive query planning:} Currently, the imputation is local to a given query, meaning
no information is shared across queries. An intriguing direction would be to take advantage
of imputation and execution data generated by repeated queries. Note that this goes beyond
simply caching imputation results, and instead could entail operations such as extending
intermediate results with prior imputation values to improve accuracy, or pruning out
query plans which have been shown to produce low-confidence imputations.
